---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: console
---



# Load the libraries
```{r}
library(readr)
library(ggplot2)
suppressPackageStartupMessages(library(dplyr))
library(forcats)
library(reshape2)
suppressPackageStartupMessages(library(choroplethr))
library(choroplethrMaps)
library(datasets)
```


# Reading the data & cleaning data

```{r}
# reading data from the csv files
team_record_df <-
  read_csv("Team Records.csv", show_col_types = FALSE)
# reading data of the statistics
cfbstats_df <- read_csv("cfbstats.csv", show_col_types = FALSE)
# reading data of the stat mapping
stat_mapping_df <-
  read_csv("Stat Mapping.csv", show_col_types = FALSE)


# Data Cleaning
# remove unwanted columns from the data
team_record_df <- team_record_df %>%
  select(-expectedWins) %>%
  select(-total.games) %>%
  select(-total.ties) %>%
  select(-conferenceGames.games) %>%
  select(-conferenceGames.wins) %>%
  select(-conferenceGames.losses) %>%
  select(-conferenceGames.ties) %>%
  select(-homeGames.games) %>%
  select(-homeGames.ties) %>%
  select(-awayGames.games) %>%
  select(-awayGames.ties)


# data cleaning of team_record_df
team_record_df <-  team_record_df %>%
  mutate(year = as.factor(year)) %>%
  mutate(team = as.factor(team)) %>%
  mutate(conference = as.factor(conference)) %>%
  mutate(total.wins = as.integer(total.wins)) %>%
  mutate(total.losses = as.integer(total.losses)) %>%
  mutate(homeGames.wins = as.integer(homeGames.wins)) %>%
  mutate(homeGames.losses = as.integer(homeGames.losses)) %>%
  mutate(awayGames.wins = as.integer(awayGames.wins)) %>%
  mutate(awayGames.losses = as.integer(awayGames.losses))


# data cleaning of cfbstats_df
cfbstats_df <- cfbstats_df %>%
  mutate(season = as.factor(season)) %>%
  mutate(team = as.factor(team)) %>%
  mutate(statName = as.factor(statName)) %>%
  mutate(conference = as.factor(conference))

# data cleaning of Stat Mapping
stat_mapping_df <- stat_mapping_df %>%
  mutate(statName = as.factor(statName)) %>%
  mutate(OffenseDefenseSpTeams = as.factor(OffenseDefenseSpTeams)) %>%
  mutate(positivenegative = as.factor(positivenegative))



# Create a mapping between team_record and cfbstats table through a
# customized lookup value

# Construct a lookup column - which is combination of year and team name
# from team_record_df
# Example: '2021 Michigan'
team_record_df <- team_record_df %>%
  mutate(cfb_lookup = paste(team_record_df$year,  team_record_df$team))

# Construct a lookup column - which is combination of season and team value
# from cfbstats_df
# Example: '2021 Michigan'
cfbstats_df <- cfbstats_df %>%
  mutate(cfb_lookup = paste(cfbstats_df$season, cfbstats_df$team))

# Merge the data frames based on cfb_lookup values
merge_team_cfb <-
  merge(cfbstats_df, team_record_df, by = "cfb_lookup")


# Now merge merge_team_cfb and stat_mapping_df
# This final data would be used for further processing
# Merge the data frames based on statName values
final_merged_df <-
  merge(merge_team_cfb, stat_mapping_df, by = "statName")


write_csv(final_merged_df, "temp2.csv")

# Splitting of the dataframe to smaller lists

# split the data to a list of dataframes - based on statName
list_by_statName <- split(final_merged_df, final_merged_df$statName)






# ******************************************************************************
# Function: append_list
#
# Purpose: The purpose of the function is to
#          extract each dataframe from the list and
#          sort the data based on statValue
#          Create a column in the dataframe - RankVal and assign the ranking values
#
# Input : list_by_statName_year_b (list of dataframes)
#
# Output: Returns the list of modified dataframes
# ******************************************************************************

# Creating a function which creates a column RankVal
# and append to the whole_list_df
append_list <- function(list_by_statName_year_b) {
  # Assign Rank Values for each year
  # based on the descending order of statValues
  for (inner_loop_idx in 1:length(list_by_statName_year_b)) {
    list_by_statName_year_b[[inner_loop_idx]] <-
      list_by_statName_year_b[[inner_loop_idx]] %>%
      mutate(RankVal =
               rank(desc(list_by_statName_year_b[[inner_loop_idx]]$statValue))) %>%
      arrange(RankVal)
  }
  
  return (list_by_statName_year_b)
}

# Create an empty list to collect all processed data
whole_list_df <- list()

# Create a loop - which goes through all entries in list based on statName
for (outer_loop_idx in 1:length(list_by_statName)) {
  # Since the list contains data for many years(2017-2021)
  # split the list further based on year
  list_by_statName_year_a <-
    split(list_by_statName[[outer_loop_idx]],
          list_by_statName[[outer_loop_idx]]$year)
  
  # Collect the return value from the function to a temporary list
  temp_list <- append_list(list_by_statName_year_a)
  
  # Collect the data from the list returned from the function 
  whole_list_df[[length(whole_list_df) + 1]] <- temp_list[[1]]
  whole_list_df[[length(whole_list_df) + 1]] <- temp_list[[2]]
  whole_list_df[[length(whole_list_df) + 1]] <- temp_list[[3]]
  whole_list_df[[length(whole_list_df) + 1]] <- temp_list[[4]]
  whole_list_df[[length(whole_list_df) + 1]] <- temp_list[[5]]
  
}

```


# Filtered Data According to Conference
```{r}

# Select only the five conferences that we are interested in
final_merged_conf_df <-
  filter(
    final_merged_df,
    conference.x == "Big Ten" |
      conference.x == "Big 12" |
      conference.x == "ACC" |
      conference.x == "Pac-12" |
      conference.x == "SEC"
  )
# select the years that we are interested in
# discard the year 2020 - since there is inconsistency in the number
# of games played by each team
final_merged_conf_df <-
  filter(
    final_merged_conf_df,
    season == "2017" |
      season == "2018" |
      season == "2019" |
      season == "2021"
  )
# split the refined dataframe according to the conferences
conference_df_sorted_list <-
  split(final_merged_conf_df, final_merged_conf_df$conference.x)

# Save the list to separate conference dataframes for plotting.
ACC_data_frame <- as.data.frame(conference_df_sorted_list["ACC"])
Big_10_data_frame <-
  as.data.frame(conference_df_sorted_list["Big Ten"])
Big_12_data_frame <-
  as.data.frame(conference_df_sorted_list["Big 12"])
Pac_12_data_frame <-
  as.data.frame(conference_df_sorted_list["Pac-12"])
SEC_data_frame <- as.data.frame(conference_df_sorted_list["SEC"])

```




# TEMP SECTION - helper statements
TODO: This section onwards - we need to select 
what should be plotted and analyzed.
```{r}
# this captures the entire list(whole_list_df) into a text document
# where we can verify the data
# since the whole_list_df is not a dataframe, we cannot export to excel
capture.output((whole_list_df), file = "temp2.txt")


# Accessing the statValue column from the 1st entry of the 
# list - which is the entry for '2017 - firstDowns'
print(whole_list_df[[1]]$statValue)

# Sample plotting of graph
# we could pick the required dataframe to 
# a separate dataframe for plotting - like this
plot_df <- whole_list_df[[1]]
write_csv(plot_df, file = "temp1.csv")
qplot(x = statValue, data = plot_df, geom = "histogram")

qplot(y = statValue, x = total.wins, data = plot_df, geom = c("point", "line"))



# or alternatively, we could directly access 
# from the qplot statements
qplot(x = whole_list_df[[1]]$statValue, 
      data = whole_list_df[[1]], geom = "histogram")


# Plot of Homegame wins and Away Games Wins
  qplot(x = awayGames.wins,
     y = homeGames.wins,
     data = final_merged_df, 
     geom = "point") + 
  geom_point(shape = "square", size = 1.5, colour = "#112446") +
  scale_fill_hue(direction = 1) +
  labs(
    x = "Away Games Wins",
    y = "Home Game Wins",
    title = "Home Game and Away Games Wins"
  ) +
  theme_minimal()


# https://www.statology.org/scatterplot-with-regression-line-r/
#create some fake data
data <- data.frame(x = c(1, 1, 2, 3, 4, 4, 5, 6, 7, 7, 8, 9, 10, 11, 11),
                   y = c(13, 14, 17, 12, 23, 24, 25, 25, 24, 28, 32, 33, 35, 40, 41))

#create scatterplot of data
plot(data$x, data$y)

model <-  lm(y ~ x, data = data)
abline(model)

#define range of x values
newx = seq(min(data$x),max(data$x),by = 1)

#find 95% confidence interval for the range of x values
conf_interval <- predict(model, newdata=data.frame(x=newx), interval="confidence",
                         level = 0.95)

#create scatterplot of values with regression line
plot(data$x, data$y)
abline(model)

#add dashed lines (lty=2) for the 95% confidence interval
lines(newx, conf_interval[,2], col="blue", lty=2)
lines(newx, conf_interval[,3], col="blue", lty=2)

```

# Filtering of Data - Jen's Method

```{r}
# converted the full list of list to data frame
all_df <- data.frame(Reduce(rbind, whole_list_df))
ACCSEC_df <- filter(all_df, conference.x == "Big Ten" )
ACCSEC_df <- ACCSEC_df %>% filter(season == "2017" |
                                    season == "2018" | 
                                    season == "2019" |
                                    season == "2021")

names(ACCSEC_df)
plot_data <- filter(
  ACCSEC_df,
  OffenseDefenseSpTeams == "Defense" &
    positivenegative == "positive"
)

# Now group the data according to team to get the summary statistics
# and prepare the summarized data
# Collect the mean/average of the statValue
# arranged by team name over the years.
plot_data <- plot_data %>% 
  group_by(team.x) %>% 
  summarise(Avg_Stat_Val = mean(statValue))

# perform plotting of the data into a horizontal column chart
# colored by team name.
# Arranged by top scoring team at the top
plot_data  %>%  mutate(team.x = fct_reorder(team.x, Avg_Stat_Val)) %>% 
ggplot( aes(y = team.x, x = Avg_Stat_Val, fill = team.x)) +
  geom_col()



```





# Monica's new code 1

```{r}
#Monica's charts
#Total statValue to total wins color coded by conference (defensive stats only)
#rename 'final_merged_df' to tmp
tmp <- final_merged_df
#filter by 2017
year <- filter(tmp, season == "2017")
#filter by Defense
defense <- filter(year, OffenseDefenseSpTeams == "Defense")
#change 'team.x' to as.factor (not sure if needed)
tmp2 <- mutate(defense, team.x = as.factor(team.x))
#group by team, total wins, and conference. Sum stats
tmp3 <- tmp2 %>% group_by(team.x, total.wins, conference.x) %>%
  summarise(across(c(statValue), sum))
#plot
p <- qplot(statValue, total.wins, data = tmp3, geom = "point", color = conference.x)

#group by conference, stats, wins
tmp4 <- tmp2 %>% group_by(conference.x) %>%
  summarise(across(c(statValue, total.wins), sum))
#plot
p2 <- qplot(statValue, total.wins, data = tmp4, geom = "point", color = conference.x)

```

# Monica's new code 2
```{r}
#Monica's charts
#Total statValue to total wins color coded by conference (defensive stats only)
#rename 'final_merged_df' to tmp
a <- final_merged_df
#filter by 2017
b <- filter(a, season == "2017")
#filter by Defense
c <- filter(b, OffenseDefenseSpTeams == "Offense")
#group by team, total wins, and conference. Sum stats
d <- c %>% group_by(team.x, total.wins, conference.x) %>%
  summarise(across(c(statValue), sum))
#plot
p3 <- qplot(statValue, total.wins, data = tmp3, geom = "point", color = conference.x)

#group by conference, stats, wins
e <- c %>% group_by(conference.x) %>%
  summarise(across(c(statValue, total.wins), sum))
#plot
qplot(statValue, total.wins, data = e, geom = "point", color = conference.x)

```

# Question: Offensive/Defensive Advantageous
```{r}
# collect the data from the final_merged_df
tmp <- final_merged_df
tmp1 <- NULL
# perform group_by and get the statistics 
tmp1 <- tmp %>% group_by(OffenseDefenseSpTeams) %>% 
  summarise(win_ct = (n()/nrow(tmp)))

# Get Ready to switch to Python
library(reticulate)
use_python("C:\\Python310")


```
## Python Implementation of pie chart
```{python}
# https://matplotlib.org/stable/gallery/pie_and_polar_charts/pie_features.html
import numpy as np
import matplotlib.pyplot as plt
# Get the data from the r terminal - those labels
labels = r.tmp1.OffenseDefenseSpTeams
# Get the corresponding values - convert to %
sizes = r.tmp1.win_ct * 100
# adjust this value for pie chart to explode separate section
explode = (0, 0.1, 0, 0) 
fig1, ax1 = plt.subplots()
# set the format in the pie chart
ax1.pie(sizes, explode=explode, labels=labels, autopct='%1.1f%%', shadow=True, startangle=90)
ax1.axis('equal') 
# show the pie chart
plt.show()

```

## R Implementation of pie chart
```{r}

x <- tmp1$win_ct
labels <- tmp1$OffenseDefenseSpTeams
piepercent<- round(100*x/sum(x), 1)
# Plot the chart.
pie(x, labels = piepercent, main = "Offensive/Defensive pie chart",col = rainbow(length(x))) 
legend("topright", legend = labels, cex = 0.8, fill = rainbow(length(x)))

# --------------------------------------------------------



# backup pie chart in R terminal
# https://r-graph-gallery.com/piechart-ggplot2.html
ggplot(tmp1,
       aes(x = "", y = win_ct, fill = OffenseDefenseSpTeams)) +
  geom_bar(stat = "identity",
           width = 1,
           color = "white") +
  coord_polar("y", start = 0) +
  theme_void() 
  


#change 'team.x' to as.factor (not sure if needed)
tmp2 <- mutate(defense, team.x = as.factor(team.x))
#group by team, total wins, and conference. Sum stats
tmp3 <- tmp2 %>% group_by(team.x, total.wins, conference.x) %>%
  summarise(across(c(statValue), sum))
#plot
p <- qplot(statValue, total.wins, data = tmp3, geom = "point", color = conference.x)


```

# Analysis of Big 10

Big 10 is strong in the defense and here in this section we are trying to prove
that over the years - they are strong in defense. 
```{r}

# ------------------------------------------------------------------
# Part 1: Average Statistics of the teams in Big 10
# ------------------------------------------------------------------
# prepare the required data for plotting the data
plot_data <- Big_10_data_frame

# Since Big-10 is strong in the defense- we select the data that is
# defense and positive from Big-10 conference.
plot_data <- filter(
  Big_10_data_frame,
  Big.Ten.OffenseDefenseSpTeams == "Defense" &
    Big.Ten.positivenegative == "positive"
)

# Now group the data according to team to get the summary statistics
# and prepare the summarized data
# Collect the mean/average of the statValue
# arranged by team name over the years.
plot_data <- plot_data %>%
  group_by(Big.Ten.team.x) %>%
  summarise(Avg_Stat_Val = mean(Big.Ten.statValue))



# perform plotting of the data into a horizontal column chart
# colored by team name.
# Arranged by top scoring team at the top
p <- plot_data  %>%  mutate(Big.Ten.team.x =
                              fct_reorder(Big.Ten.team.x, Avg_Stat_Val)) %>%
  ggplot(aes(y = Big.Ten.team.x, x = Avg_Stat_Val, fill = Big.Ten.team.x)) +
  geom_col() +  
  labs(title = "Plot of Big Ten Teams with their Average statValues",
       x = "Average Statistic Values",
       y = "Big 10 Teams",
       fill = "Big 10 Teams")



# Create a common theme for all plots
theme_common <- theme(
  plot.title = element_text(size = 20, hjust = 0.5, margin = margin(
      b = -0.1, t = 0.4, l = 2, unit = "cm") ),
  legend.title = element_text(hjust = 0.4 , vjust = 0.3, size = 10),
  plot.background = element_rect(fill = "#f5f5f2", color = NA),
  panel.background = element_rect(fill = "#f5f5f2", color = NA),
  legend.background = element_rect(fill = "#f5f5f2", color = NA)
)

p + theme_common
# ------------------------------------------------------------------
# Part 2: Display of the team statistics in the form of map
# ------------------------------------------------------------------
#
# reinitialize the dataframe to prepare for the map data
# Collect the required values from the Big_10_data_frame
# summarize the values by teams
plot_data <- NULL
plot_data <- filter(
  Big_10_data_frame,
  Big.Ten.OffenseDefenseSpTeams == "Defense" &
    Big.Ten.positivenegative == "positive"
) %>%
  group_by(Big.Ten.team.x) %>%
  summarise(value = mean(Big.Ten.statValue))

# Create a separate column with the heading as
# region to get the states to map
plot_data <- plot_data %>%
  mutate(region = Big.Ten.team.x) %>%
  arrange(desc(value))


# cleanup the state names(region) based on the data collected data
plot_data <- mutate(plot_data, region = fct_recode(region, Ohio = "Ohio State"))
plot_data <- mutate(plot_data, region = fct_recode(region, Pennsylvania = "Penn State"))
plot_data <- mutate(plot_data, region = fct_recode(region, Michigan = "Michigan State"))
plot_data <- mutate(plot_data, region = fct_recode(region, Indiana = "Purdue"))
plot_data <- mutate(plot_data, region = fct_recode(region, Illinois = "Northwestern"))
plot_data <- mutate(plot_data, region = fct_recode(region, `New Jersey` = "Rutgers"))


# convert the state names to lower cases to enable plotting
# Summarize the data to be consolidated based on states instead of universities.
plot_data <- plot_data %>% group_by(region) %>% 
  summarise(value = sum(value)) %>% 
  mutate(region = tolower(region))


# show the list of states that are involved in the Big 10 Conferences
# Creating a different visualization of average statValues across Big 10 Conferences
# Apply the common theme
# choropleth map zooming to the specific states that we are interested in
state_choropleth(plot_data, 
                 zoom = c("ohio","wisconsin",
                          "iowa","pennsylvania",
                          "michigan","indiana",
                          "illinois","minnesota",
                          "nebraska",
                          "maryland","new jersey"),
                 title = "Big 10 Conference States",
                 legend = "Average StatValue") +
  theme_common

# ------------------------------------------------------------------
# Part 3: Trend of the teams over the years
# ------------------------------------------------------------------
# compare the average statistics for each team over the years - to get 
# the historical raise and fall of the teams.

plot_data <- NULL
# prepare the required data for plotting the data
plot_data <- Big_10_data_frame

# Since Big-10 is strong in the defense- we select the data that is
# defense and positive from Big-10 conference.
plot_data <- filter(
  Big_10_data_frame,
  Big.Ten.OffenseDefenseSpTeams == "Defense" &
    Big.Ten.positivenegative == "positive"
)

# Now group the data according to team to get the summary statistics
# and prepare the summarized data
# Collect the mean/average of the statValue
# arranged by team name over the years.
plot_data <- plot_data %>%
  group_by(Big.Ten.team.x, Big.Ten.year) %>%
  summarise(Avg_Stat_Val = mean(Big.Ten.statValue))


# plot the data for each team over the years - so that we
# know the trend about the teams
p <- ggplot(data = plot_data, 
         aes(x = Big.Ten.team.x, y = Avg_Stat_Val,  stat = "identity", fill = Big.Ten.year)) + 
       geom_bar(stat = "identity", position = position_dodge(), alpha = 0.75) +
    geom_text(aes(label = round(Avg_Stat_Val)), vjust = 1.5,
              position = position_dodge(width = .9), size = 3, color = "black", fontface = "italic") +
    facet_grid(cols = vars(Big.Ten.year)) + 
    labs(title = "Plot of Big Ten Teams over the years",
       x = "Average Statistic Values",
       y = "Big 10 Teams",
       fill = "Years") + 
    scale_fill_brewer(palette = "Dark2") + 
    theme(plot.title = element_text(size = 20, face = "bold.italic"),
      axis.text.x = element_text(angle = 90, face = "bold.italic", size = 10, vjust = 0.75, hjust = 1), 
          axis.title = element_text(size = 14, face = "italic"),
          axis.text.y = element_text(face = "bold.italic", size = 10),
          panel.grid.minor = element_blank(),
          panel.grid.major = element_blank())

# Save the plot to a file
ggsave(filename = "Indi_Years.png", plot = p, width = 10, height = 6, units = "in")

# perform plotting of the data into a horizontal column chart
# colored by team name.
# Arranged by top scoring team at the top
plot_data  %>%  mutate(Big.Ten.team.x =
                              fct_reorder(Big.Ten.team.x, Avg_Stat_Val)) %>%
  ggplot(aes(y = Big.Ten.team.x, x = Avg_Stat_Val, fill = Big.Ten.team.x)) +
  geom_col() +
  labs(title = "Plot of Big Ten Teams with their Average statValues",
       x = "Average Statistic Values",
       y = "Big 10 Teams",
       fill = "Big 10 Teams")



```

